Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-06-09 doc/thesis/abstract.tex
Edits and edits

Finally, the Barab\'{a}si-Albert model was fit to six real world HIV datasets,
and substantial heterogeneity in the parameter estimates was observed.
{\color{red}\sout{Point estimates}}{\color{blue}\uline{Depending on the choice
of prior, posterior means}} for the preferential attachment power ranged from
{\color{red}\sout{0.06 to 1.05}} {\color{blue}\uline{0.26, slightly less than
logarithmic, to 1.00, exactly linear.}} {\color{blue}\uline{Point estimates of
the strength of preferential attachment were higher in two injection drug user
populations, potentially indicating that high-degree ''hub'' nodes may play a
role in epidemics among this risk group.}} Our results underscore the
importance of considering contact structures when {\color{red}\sout{performing
phylodynamic inference}}{\color{blue}\uline{investigating viral outbreaks.}}

>>>>

Finally, the Barab\'{a}si-Albert model was fit to six real world HIV datasets,
and substantial heterogeneity in the parameter estimates was observed.
{\color{red}\sout{Point estimates}}{\color{blue}\uline{Depending on the choice
of prior, posterior means}} for the preferential attachment power ranged from
{\color{red}\sout{0.06 to 1.05}} {\color{blue}\uline{0.26, slightly less than
logarithmic, to 1.00, exactly linear.}} {\color{blue}\uline{Point estimates of
the strength of preferential attachment were higher in injection drug user
populations, potentially indicating that high-degree ``hub'' nodes may play a
role in epidemics among this risk group.}} Our results underscore the
importance of considering contact structures when {\color{red}\sout{performing
phylodynamic inference}}{\color{blue}\uline{investigating viral outbreaks.}}

### 2016-06-09 doc/thesis/intro.tex
Edits and edits

{\color{red}\sout{Ordinarily, Bayesian inference targets the posterior
distribution $\Pr(\theta \mid y)$. That is, in the Bayesian framework,}} 
{\color{blue}\uline{By targeting the posterior distribution, Bayesian inference
makes the assertion that}} model parameters with higher posterior density are
``better'' in the sense that they offer a more credible explanation for the
observed data. The approximate posterior targeted by \gls{ABC} uses an
alternative metric for parameter credibility, namely the similarity of
simulated datasets to the observed data. If datasets simulated under the model
closely resemble the real data, it follows that the model is a reasonable
approximation to the real-world process generating the observed data. More
formally, {\color{blue}\uline{let $y$ be the observed data to which we are
trying to fit a model with parameters $\theta$. In the case of this work, the
data is a transmission tree $T$, but we shall stick with the generic variable
$y$ for now.}} Suppose we have a distance measure $\rho$ defined on the space
of all possible data our model could generate. \gls{ABC} aims to sample from
the joint posterior distribution of model parameters and simulated datasets $z$
which are within some small distance $\varepsilon$ of the observed data $y$,
\[
  \pi_{\varepsilon}(\theta, z \mid y) =
  \frac{\Pr(\theta) \Pr(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
  {\int_{A_{\varepsilon, y} \times \Theta} \Pr(\theta) \Pr(z \mid \theta) \d \theta}.
\]
Here, $A_{\varepsilon, y}$ is an $\varepsilon$-ball around $y$ with
respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
$\I$ is the indicator function~\autocite{marin2012approximate}.
{\color{blue}\uline{The term $\Pr(z \mid \theta)$ appears to be the bothersome
likelihood again, but this will turn out not to be a problem because we are
simulating $z$ ourselves. In fact, it is possible to sample from this
distribution exactly (see next section).}}

@@

{\color{red}\sout{As we shall see in the next section, this distribution can
be sampled from exactly. The word ``approximate'' derives from the assumption
that,}}
{\color{blue}\uline{It is not yet obvious that the \gls{ABC} target
distribution just presented can in fact be used to approximate the true
posterior. TODO }}

for a suitably chosen distance $\rho$ and a small enough $\varepsilon$, the
marginal in $z$ of this distribution approximates the posterior of
interest~\autocite{marin2012approximate}. That is,
\[
  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx f(\theta \mid y).
\]
The distribution $\pi_\varepsilon(\theta, z \mid y)$ is variously referred to
as the \textit{\gls{ABC} target distribution} or the \gls{ABC} approximation
to the posterior. The intuition for why this approximation might be reasonable
comes from the fact that, when $\varepsilon = 0$, the $\varepsilon$-ball around
$y$ should contain only $y$ itself, hence the integral on the left is exactly
equal to the posterior. Thus, by taking $\varepsilon$ small, we should attain
something close to the posterior {\color{blue}\uline{if $\rho$ captures the
similarity between datasets reasonably well. However, the accuracy of the
\gls{ABC} approximation depends heavily on the choice of distance function
\autocite{aeschbacher2012novel,blum2013comparative}.}}

In many applications {\color{blue}\uline{(eg. \autocite{fu1997estimating,
tanaka2006using})}}, $\rho$ is defined as $\rho(S(\cdot), S(\cdot))$ where $S$
is a function which maps data points into a vector of summary statistics.
{\color{blue}\uline{In the context of \gls{ABC}, a summary statistic $S$ is
called \defn{sufficient} if} 
\[
  f(\theta \mid y) = f(\theta \mid S(y)).
\]
\uline{That is, sufficiency implies that the data can be replaced with the
summary statistic without losing any information about the posterior
distribution~\autocite{marjoram2006modern}. For most problems, it is not
possible to find sufficient summary statistics~\autocite{marjoram2006modern}.
A number of sophisticated methods have been developed for selecting and
weighting summary statistics based on various optimality criteria}
\autocite[][and references therein]{aeschbacher2012novel,
blum2013comparative}}.

@@

The most recently developed class of algorithm for \gls{ABC} is
\gls{ABC}-\gls{SMC}~\autocite{sisson2007sequential, beaumont2009adaptive}. As
with \gls{ABC}-\gls{MCMC}, the algorithm is a straightforward modification of
an existing Bayesian inference method, in this case the \gls{SMC} sampler
(\cref{subsec:smcsamp}). The sequence of target distributions is defined as
$\pi_i = \pi_{\varepsilon_i}(x \mid y)$ for a decreasing sequence of tolerances
$\varepsilon_i$. The intention is for the algorithm to progress smoothly
through a sequence of target distributions which ends at the \gls{ABC}
approximation to the posterior. {\color{blue}\uline{The initial importance
distribution, denoted as $\eta_0$ in} \cref{alg:smcsamp}, \uline{is taken to be}
$\Pr(\theta) \Pr(z \mid \theta)$. TODO}

@@

\begin{algorithm}
  \caption{\gls{ABC}-\gls{SMC}.}
  \begin{algorithmic}
    \For {$k = 1$ to $n$}
      \State Sample $x_1^{(k)}$ from $\eta_1(x_1)$
      \Comment{Initialize the $k$th particle}
      \State $w^{(k)} \gets \gamma_1(x_1^{(k)}) / \eta_1(x_1^{(k)})$
      \State Normalize the weights so that $\sum w = 1$
    \EndFor
    \For {$i = 2$ to $d$}
      \For {$k = 1$ to $n$}
        \State Sample $x_i^{(k)}$ from $K(x_{i-1}^{(k)}, x_i)$
        \Comment Extend the $k$th particle
        \State $w^{(k)} \gets w^{(k)} \cdot \dfrac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}{\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}$
      \EndFor
      \State Normalize the weights so that $\sum w = 1$
      \If {$\ESS(w) < T$}
        \Comment{$T$ is a user-defined threshold}
        \State Resample the particles according to $w$
        \For {$k = 1$ to $n$}
          \State $w^{(k)} \gets 1/n$
        \EndFor
      \EndIf
    \EndFor
  \end{algorithmic}
  \label{alg:abcsmc}
\end{algorithm}

>>>>

{\color{red}\sout{Ordinarily, Bayesian inference targets the posterior
distribution $\Pr(\theta \mid y)$. That is, in the Bayesian framework,}} 
{\color{blue}\uline{By targeting the posterior distribution, Bayesian inference
makes the assertion that}} model parameters with higher posterior density are
``better'' in the sense that they offer a more credible explanation for the
observed data. The approximate posterior targeted by \gls{ABC} uses an
alternative metric for parameter credibility, namely the similarity of
simulated datasets to the observed data. If datasets simulated under the model
closely resemble the real data, it follows that the model is a reasonable
approximation to the real-world process generating the observed data. More
formally, {\color{blue}\uline{let $y$ be the observed data to which we are
trying to fit a model with parameters $\theta$. In the case of this work, the
data is a transmission tree $T$, but we shall stick with the generic variable
$y$ for now.}} Suppose we have a distance measure $\rho$ defined on the space
of all possible data our model could generate. \gls{ABC} aims to sample from
the joint posterior distribution of model parameters and simulated datasets $z$
which are within some small distance $\varepsilon$ of the observed data $y$,
\begin{align}
  \pi_{\varepsilon}(\theta, z \mid y) =
  \frac{\Pr(\theta) \Pr(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
  {\int_{A_{\varepsilon, y} \times \Theta} \Pr(\theta) \Pr(z \mid \theta) \d \theta}.
  \label{eq:abctarget}
\end{align}
Here, $A_{\varepsilon, y}$ is an $\varepsilon$-ball around $y$ with
respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
$\I$ is the indicator function~\autocite{marin2012approximate}. The
distribution $\pi_\varepsilon(\theta, z \mid y)$ will be referred to as
the \textit{\gls{ABC} target distribution}. {\color{blue}\uline{The term $\Pr(z
\mid \theta)$ appears to be the bothersome likelihood again, but this will turn
out not to be a problem because we are simulating $z$ ourselves. In fact, it is
possible to sample from this distribution exactly (see next section).}}

@@

{\color{red}\sout{As we shall see in the next section, this distribution can
be sampled from exactly. The word ``approximate'' derives from assumption that,
for a suitably chosen distance $\rho$ and a small enough $\varepsilon$, the
marginal in $z$ of this distribution approximates the posterior of
interest~\autocite{marin2012approximate}. That is, }}
{\color{blue}\uline{In fact, it is not the \gls{ABC} target distribution
itself, but rather its marginal in $z$, which approximates the posterior
distribution. In other words, we claim that }} 
\[
  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx \Pr(\theta \mid y).
\]
The intuition for why this approximation might be reasonable
comes from the fact that, when $\varepsilon = 0$, the $\varepsilon$-ball around
$y$ should contain only $y$ itself, hence the integral on the left is exactly
equal to the posterior. Thus, by taking $\varepsilon$ small, we should attain
something close to the posterior {\color{blue}\uline{if $\rho$ captures the
similarity between datasets reasonably well. However, the accuracy of the
\gls{ABC} approximation depends heavily on the choice of distance function
\autocite{aeschbacher2012novel,blum2013comparative}.}}

\subsubsection{Distance functions and summary statistics in ABC}

In many applications {\color{blue}\uline{(eg. \autocite{fu1997estimating,
tanaka2006using})}}, $\rho$ is defined as $\rho(S(\cdot), S(\cdot))$ where $S$
is a function which maps data points into a vector of summary statistics.
{\color{blue}\uline{In the context of \gls{ABC}, a summary statistic $S$ is
called \defn{sufficient} if} 
\[
  \Pr(\theta \mid y) = \Pr(\theta \mid S(y)).
\]
\uline{That is, sufficiency implies that the data can be replaced with the
summary statistic without losing any information about the posterior
distribution~\autocite{marjoram2006modern}. For most problems, it is not
possible to find sufficient summary statistics~\autocite{marjoram2006modern}.
A number of sophisticated methods have been developed for selecting and
weighting summary statistics based on various optimality criteria}
\autocite[][and references therein]{aeschbacher2012novel,
blum2013comparative}}.

@@

The most recently developed class of algorithm for \gls{ABC} is
\gls{ABC}-\gls{SMC}~\autocite{sisson2007sequential, beaumont2009adaptive}. As
with \gls{ABC}-\gls{MCMC}, the algorithm is a straightforward modification of
an existing Bayesian inference method, in this case the \gls{SMC} sampler
(\cref{subsec:smcsamp}). The sequence of target distributions is defined as
$\pi_i(x) = \pi_{\varepsilon_i}(x \mid y)$ for a decreasing sequence of
tolerances $\varepsilon_i$. The intention is for the algorithm to progress
smoothly through a sequence of target distributions which ends at the \gls{ABC}
approximation to the posterior. {\color{blue}\uline{The initial value
$\varepsilon_0$ is set to $\infty$, which makes the first distribution in the
sequence}
\[
  \pi_0(\theta, z) = \frac{\Pr(\theta) \Pr(z \mid \theta)}
               {\int_{\R \times \Theta} \Pr(\theta) \Pr(z \mid \theta) \d\theta \d z}.
\]
\uline{For brevity, we have ommitted the dependence on the observed data $y$.
In the terminology of} \cref{alg:smc}, \uline{the numerator is the first of the
$\gamma$'s, that is, $\gamma_0 = \Pr(\theta) \Pr(z \mid \theta)$. Sampling in
proportion to $\gamma_0$ is straightforward and was already demonstrated for
rejection \gls{ABC} above. Because the sampling is exact, the initial
importance weights are all set equal to 1 and normalized to $1/n$ where $n$ is
the number of particles.}}

@@

\begin{algorithm}
  \caption{\gls{ABC}-\gls{SMC}.}
  \begin{algorithmic}
    \For {$k = 1$ to $n$}
      \State Sample $x_1^{(k)}$ from $\Pr(\theta) \Pr(z \mid \theta)$
      \Comment{Draw $\theta$ from the prior and simulate $z$ from $\theta$}
      \State $w^{(k)} \gets 1/n$
    \EndFor
    \For {$i = 2$ to $i_{\max}$}
      \Comment{$i_{\max}$ is a user-specified number of iterations}
      \For {$k = 1$ to $n$}
        \State Sample $x_i^{(k)}$ from $K(x_{i-1}^{(k)}, x_i)$
        \Comment Extend the $k$th particle
        \State $w^{(k)} \gets w^{(k)} \cdot \dfrac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}{\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}$
      \EndFor
      \State Normalize the weights so that $\sum w = 1$
      \If {$\ESS(w) < T$}
        \Comment{$T$ is a user-defined threshold}
        \State Resample the particles according to $w$
        \For {$k = 1$ to $n$}
          \State $w^{(k)} \gets 1/n$
        \EndFor
      \EndIf
    \EndFor
  \end{algorithmic}
  \label{alg:abcsmc}
\end{algorithm}

### 2016-06-09 doc/thesis/methods_results.Rnw
Edits and edits

{\color{blue}\uline{ In this chapter, we will address the three research aims
of this thesis introduced in} \cref{sec:obj}. \uline{First, in}
\cref{sec:netabc}, \uline{we describe \software{netabc}, a computer program
that implements an \acrlong{ABC}-based algorithm to fit contact network models
to phylogenetic data. We also provide a justification for the use of \gls{ABC}
for this problem by arguing that the likelihood functions required to fit these
models by more conventional means are likely to be computationally intractable.
Second, in} \cref{sec:ba}, \uline{we perform a simulation study to investigate
the \acrlong{BA} network model, which uses a preferential attachment mechanism
to generate networks with the power law degree distributions observed in real
world social and sexual networks. We progress through two exploratory analyses
testing the identifiability of the model's parameters, and conclude by testing
\software{netabc}'s ability to recover the parameters from simulated
transmission trees. Third, in} \cref{sec:hiv}, \uline{we apply
\software{netabc} to fit the \gls{BA} model to six real world \gls{HIV}
datasets, with the understanding of the model parameters' identifiability
gained through the simulation experiments. We conclude the chapter with a
unified discussion of the three research aims, including interpretation of the
results of both the simulated and real data experiments, as well as an
examination of the limitations of our approach and opportunities for future
investigation. }}

@@

\Cref{fig:gridalpha,fig:gridI,fig:gridm,fig:gridN} \uline{show kernel score
distributions of kernel scores along the grid for each parameter. The
distributions for some values, such as $\gls{alpha} = 1.25$, $\gls{I} = 500$
and $4500$, $\gls{m} = 1$, and $\gls{N} = 1000$, exhibited distinct peaks
around the true value. This indicates that these values produce distinctively
shaped trees that can be identified with the tree kernel, when the other
parameter values are fixed and known. However, for the majority of values of
each parameter, the score distributions were fairly flat around the true value. 
This means there is a range of values which produce similarly shaped trees, and
the parameter is less identifiable within that range.}}

@@

We downloaded all sequences associated with each published study from GenBank.
For the \textcite{novitsky2014impact} data, each \textit{env} sequence was
aligned pairwise to the HXB2 reference sequence (GenBank accession number
K03455), and the hypervariable regions were clipped out with
\software{BioPython} version 1.66+~\autocite{cock2009biopython}. Sequences were
multiply aligned using \software{MUSCLE} version 3.8.31
\autocite{edgar2004muscle}, and alignments were manually inspected with
\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
constructed from the nucleotide alignments by approximate maximum likelihood
using \software{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
\gls{GTR} model~\autocite{tavare1986some}. Transmission trees were estimated by
rooting and time-scaling the phylogenies by root-to-tip regression, using a
modified version of Path-O-Gen (distributed as part of
BEAST~\autocite{drummond2007beast}) as described
previously~\autocite{poon2015phylodynamic}.

@@

We applied kernel-assisted ABC to five published HIV datasets (\cref{tab:data}),
and found substantial heterogeneity among the parameter estimates
(\cref{fig:abchpd,fig:abchpdm2}). Plots of the marginal posterior distributions
for each dataset are shown in
\cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
estimated $\alpha$ values near unity for the prior allowing $m = 1$ (posterior
mean [95\% \gls{HPD}] 
  \Sexpr{pmean1["niculescu2015", round(alpha, 2)]} 
  [\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
   \Sexpr{up1["niculescu2015", round(alpha, 2)]}]
and
  \Sexpr{pmean1["wang2015", round(alpha, 2)]} 
  [\Sexpr{low1["wang2015", round(alpha, 2)]} -
   \Sexpr{up1["wang2015", round(alpha, 2)]}] respectively).
The estimates did not change appreciably when $m = 1$ was disallowed by the
prior, although the credible interval of the \textcite{niculescu2015recent}
data was narrower
  (\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
   \Sexpr{up1["niculescu2015", round(alpha, 2)]}).
When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
low estimated $\alpha$ values
  (\Sexpr{pmean1["li2015", round(alpha, 2)]} 
  [\Sexpr{low1["li2015", round(alpha, 2)]} - 
  \Sexpr{up1["li2015", round(alpha, 2)]}]
and
  \Sexpr{pmean1["cuevas2009", round(alpha, 2)]} 
  [\Sexpr{low1["cuevas2009", round(alpha, 2)]} -
   \Sexpr{up1["cuevas2009", round(alpha, 2)]}]). 
However, the estimates increased when $m = 1$ was not permitted, although the
HPD intervals remained roughly the same
  (\Sexpr{pmean2["li2015", round(alpha, 2)]} 
  [\Sexpr{low2["li2015", round(alpha, 2)]} - 
  \Sexpr{up2["li2015", round(alpha, 2)]}]
and
  \Sexpr{pmean2["cuevas2009", round(alpha, 2)]} 
  [\Sexpr{low2["cuevas2009", round(alpha, 2)]} -
   \Sexpr{up2["cuevas2009", round(alpha, 2)]}]).
The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
for both priors on $m$
  (\Sexpr{pmean1["novitsky2014", round(alpha, 2)]} for $m \geq 1$;
   \Sexpr{pmean2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
However, the confidence interval was much wider when $m = 1$ was allowed
  ([\Sexpr{low1["novitsky2014", round(alpha, 2)]} -
    \Sexpr{up1["novitsky2014", round(alpha, 2)]}] for $m \geq 1$ vs.
   [\Sexpr{low2["novitsky2014", round(alpha, 2)]} -
    \Sexpr{up2["novitsky2014", round(alpha, 2)]}] for $m \geq 2$).

>>>>

{\color{blue}\uline{ In this chapter, we will address the three research aims
of this thesis introduced in} \cref{sec:obj}. \uline{First, in}
\cref{sec:netabc}, \uline{we describe \software{netabc}, a computer program
that implements an \acrlong{ABC}-based algorithm to fit contact network models
to phylogenetic data. We also provide a justification for the use of \gls{ABC}
for this problem by arguing that the likelihood functions required to fit these
models by more conventional means are likely to be computationally intractable.
Second, in} \cref{sec:ba}, \uline{we perform a simulation study to investigate
the \acrlong{BA} network model, which uses a preferential attachment mechanism
to generate networks with the power law degree distributions observed in real
world social and sexual networks. We progress through two exploratory analyses
testing the identifiability of the model's parameters, and conclude by testing
\software{netabc}'s ability to recover the parameters from simulated
transmission trees. Third, in} \cref{sec:hiv}, \uline{we apply
\software{netabc} to fit the \gls{BA} model to six real world \gls{HIV}
datasets, with the understanding of the model parameters' identifiability
gained through the simulation experiments. We conclude the chapter with a
unified discussion of the three research aims, including interpretation of the
results of both the simulated and real data experiments, as well as an
examination of the limitations of our approach and opportunities for future
investigation.}}

@@

\Cref{fig:gridalpha,fig:gridI,fig:gridm,fig:gridN} \uline{show kernel score
distributions of kernel scores along the grid for each parameter. The
distributions for some values, such as $\gls{alpha} = 1.25$, $\gls{I} = 500$
and $4500$, $\gls{m} = 1$, and $\gls{N} = 1000$, exhibited distinct peaks
around the true value. This indicates that these values produce distinctively
shaped trees that can be identified with the tree kernel, when the other
parameter values are fixed and known. However, for the majority of values of
each parameter, the score distributions were fairly flat around the true value. 
This means there is a range of values which produce similarly shaped trees, and
the parameter is less identifiable within that range. The exception was
\gls{I}, whose score distributions exhibited a more or less rounded shape with
the highest point near the true value. }}

@@

We downloaded all sequences associated with each published study from GenBank.
{\color{blue}\uline{For the \textcite{niculescu2015recent} data, only sequences
from \gls{IDU} (whose sequence identifiers included the letters ``DU'') were
included in the analysis. \textcite{kao2011surveillance} found a very strong
association in their study population between subtype and risk group - subtype
B was almost always associated with \gls{MSM}, whereas \gls{IDU} were usually
infected with a circulating recombinant form. Since there were many more
subtype B sequences in their data than sequences of other subtypes, we
restricted our analysis to the subtype B sequences and labelled this dataset as
\gls{MSM}. Two datasets (\textcite{zetterberg2004two} and
\textcite{grabowski2014role}) included both \textit{env} and \textit{gag}
sequences. Each gene was analyzed separately to assess the robustness of
\software{netabc} to the particular \gls{HIV} gene sequence used to estimate a
transmission tree.}}

{\color{red}\sout{For the \textcite{novitsky2014impact} data,}} Each
\textit{env} sequence was aligned pairwise to the HXB2 reference sequence
(GenBank accession number K03455), and the hypervariable regions were clipped
out with \software{BioPython} version 1.66+~\autocite{cock2009biopython}.
Sequences were multiply aligned using \software{MUSCLE} version 3.8.31
\autocite{edgar2004muscle}, and alignments were manually inspected with
\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
constructed from the nucleotide alignments by approximate maximum likelihood
using \software{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
\gls{GTR} model~\autocite{tavare1986some}. Transmission trees were estimated by
rooting and time-scaling the phylogenies by root-to-tip regression, using a
modified version of Path-O-Gen (distributed as part of
BEAST~\autocite{drummond2007beast}) as described
previously~\autocite{poon2015phylodynamic}.

@@

We applied kernel-assisted ABC to {\color{red}\sout{five}}
{\color{blue}\uline{seven}} published HIV datasets, {\color{blue}\uline{plus
one in-house, unpublished dataset}} (\cref{tab:data}).
{\color{red}\sout{and found substantial heterogeneity among the parameter
estimates}} {\color{blue}\uline{Posterior mean point estimates and 50\% and
95\% \glspl{HPD} for each parameter are shown in} \cref{fig:abchpd}, \uline{and
marginal posterior distributions in}
\cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
\Cref{fig:abchpdm2} \uline{shows point estimates and \glspl{HPD} obtained when
the value $\gls{m} = 1$ was disallowed by the prior. Since the results
indicated that $\gls{m} = 1$ was the most credible value for several datasets, 
all results discussed henceforth are for the prior $\gls{m} \sim
\DiscreteUniform(1, 5)$ unless otherwise stated.}}

Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
estimated $\alpha$ values near unity for the prior allowing $m = 1$ (posterior
mean [95\% \gls{HPD}] 
  \Sexpr{pmean1["niculescu2015", round(alpha, 2)]} 
  [\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
   \Sexpr{up1["niculescu2015", round(alpha, 2)]}]
and
  \Sexpr{pmean1["wang2015", round(alpha, 2)]} 
  [\Sexpr{low1["wang2015", round(alpha, 2)]} -
   \Sexpr{up1["wang2015", round(alpha, 2)]}] respectively).
The estimates did not change appreciably when $m = 1$ was disallowed by the
prior, although the credible interval of the \textcite{niculescu2015recent}
data was narrower
  (\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
   \Sexpr{up1["niculescu2015", round(alpha, 2)]}).
When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
low estimated $\alpha$ values
  (\Sexpr{pmean1["li2015", round(alpha, 2)]} 
  [\Sexpr{low1["li2015", round(alpha, 2)]} - 
  \Sexpr{up1["li2015", round(alpha, 2)]}]
and
  \Sexpr{pmean1["cuevas2009", round(alpha, 2)]} 
  [\Sexpr{low1["cuevas2009", round(alpha, 2)]} -
   \Sexpr{up1["cuevas2009", round(alpha, 2)]}]). 
However, the estimates increased when $m = 1$ was not permitted, although the
HPD intervals remained roughly the same
  (\Sexpr{pmean2["li2015", round(alpha, 2)]} 
  [\Sexpr{low2["li2015", round(alpha, 2)]} - 
  \Sexpr{up2["li2015", round(alpha, 2)]}]
and
  \Sexpr{pmean2["cuevas2009", round(alpha, 2)]} 
  [\Sexpr{low2["cuevas2009", round(alpha, 2)]} -
   \Sexpr{up2["cuevas2009", round(alpha, 2)]}]).
The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
for both priors on $m$
  (\Sexpr{pmean1["novitsky2014", round(alpha, 2)]} for $m \geq 1$;
   \Sexpr{pmean2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
However, the confidence interval was much wider when $m = 1$ was allowed
  ([\Sexpr{low1["novitsky2014", round(alpha, 2)]} -
    \Sexpr{up1["novitsky2014", round(alpha, 2)]}] for $m \geq 1$ vs.
   [\Sexpr{low2["novitsky2014", round(alpha, 2)]} -
    \Sexpr{up2["novitsky2014", round(alpha, 2)]}] for $m \geq 2$).
